Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
added stuff

\newcommand{\kernelmerge}{\textsc{KernelMerge}}

@@

Note that the above guidelines are not hard-and-fast rules.  For
example, a smart scheduler might respond selectively to $\offerkill$
and $\offerfork$ calls by predicting future high-priority workloads
based on past workloads.

@@

For ease of implementation, our current prototype supports
multitasking a single cooperative kernel with a single non-cooperative
kernel (though the non-cooperative kernel can be invoked many times).
Rather than adding language support for $\transmit$ annotations, we
assume that all variables declared at the root lexical scope of a
cooperative kernel are $\transmit$-annotated; this allows our
implementation to piggy-back on top of Clang (see below) without
requiring front-end changes.  Finally, we require that $\offerkill$,
$\offerfork$ and $\resizingglobalbarrier$ are called from the entry
function of a cooperative kernel.  This allows us to use
$\keyword{goto}$ to direct threads joining the cooperative kernel to
the appropriate location, and $\keyword{return}$ to force threads to
leave the computation.  These restrictions still allow us to
experiment with interesting irregular algorithms (see
\mysec\ref{sec:experiments}), and could all be removed with
engineering effort, but this effort might not be well invested since a
full implementation of coopeartive kernels would not use the
megakernel approach we present here.

@@

We have built a tool, \kernelmerge{}, that takes a cooperative kernel
and a non-cooperative kernel and merges them to produce a megakernel
with an embedded GPU scheduler.  We implemented \kernelmerge{} using the Clang LibTooling framework,\footnote{\url{http://clang.llvm.org/docs/LibTooling.html}}
leveraging recent Clang support for OpenCL 2.0~\cite{ClangOpenCL20}.

@@

\paragraph{Efficient implementation}

The interesting interactions with respect to performance and safety
occur in the interaction of the scheduler and cooperative kernel. We
discuss these details now.

At the core of the scheduler is the workgroup pool, which maps each
workgroup to a status and keeps track of the number of workgroups in
each status. The workgroup pool can change in two ways: (1)
\emph{assigning tasks}, where workgroups are changed from available to
computing a kernel (2) \emph{becoming available}, where a workgroup
exits a kernel computation (either through completion or $\offerkill$)
and changes its status from computing a kernel to available.

\emph{Assigning tasks} occurs either when the scheduler is launches a
new kernel, or when a workgroup calls $\offerfork$ and observes that
there are available resources. This creates a potential conflict
between forking workgroups as well as the scheduler, as they may try
to assign the same available workgroup to different tasks
simultaneously. To fix this conflict, we introduce an
\emph{assignment} mutex in the scheduler context.

When the scheduler determines a new kernel has been launched, it spins
waiting to acquire the assignment lock. Once acquired, it determines
how to obtain the required number of workgroups (possibly by
signalling a cooperative kernel to enable $\offerkill$). The scheduler
than waits until the required number of workgroups become available.
Because it holds the assignment mutex, the scheduler knows that the
number of available workgroups in the workgroup pool can only
increase.

When $\offerfork$ is executed, the calling workgroup has a single
attempt to acquire the assignment mutex.  If it fails, the call
returns without a fork. This provides a quick return of $\offerfork$
when another actor is doing an assigning task. If the assignment mutex
is acquired, then the calling workgroup can check the scheduler
context to see if its allowed to fork by the scheduler and safely
assign available workgroups from the workgroup pool in isolation from
other actors that which to assign tasks. \TSComment{it also updates
  the kernel context, and sends transmit variables while holding this
  lock}

As opposed to assignment tasks, availability tasks can be performed in
a lock-free manner without conflict. This is because a workgroup only
modifies its own state in the task pool and can atomically increment
and decrement the required counters. Availability tasks are also able
to be performed lock-free with assignment tasks (although accesses to
the workgroup state and counters must be performed
atomically). Availability tasks will only increase the number of
available workgroups, and do not conflict with assignment tasks
sending available workgroups to kernels.

Although assignment tasks to the workgroup pool can be performed
concurrently, concurrent calls to $\offerkill$ must be protected
because workgroups that exit (a) modify the kernel context which
causes a conflict and (b) the semantics of $\offerkill$ guarantee that
only the workgroup with the highest id can exit. Rather than introduce
a new mutex to protect $\offerkill$, we instead use the number of
active groups from the kernel context.  The call to $\offerkill$
begins by checking the calling workgroup id against the number of
active groups for the kernel. If the calling workgroup is the last
workgroup, it enters and is then able to check if it can be
killed. Otherwise it immediately returns. If a workgroup determines
it should be killed, it updates

From this implementation, it falls out that $\offerkill$ and
$\offerfork$ from the same workgroup can never succeed concurrently.
If $\offerkill$ terminates the calling workgroup, then the scheduler
is holding the assignment lock and will continue to hold it until 
the call to $\offerkill$ finishes. Thus, any call to $\offerfork$
during this time will not result in forking workgroups.

@@

\ADComment{Optimised barrier implementation.}

@@

Our experiments span 13 pairs of cooperative kernels and data sets, 3
classes of non-cooperative kernel workloads and frequency, 4 quantities
of workgroups claimed for the non-cooperative kernel and 2 variations of
resizing barriers for graph algorithms, leading to 288
configurations. Each of this configuration is run 10 times on each
GPU. For each run, we record the execution time of the cooperative
kernel. For each scheduling of the non-cooperative kernel during the
run, we also record the \emph{gather time} needed by the scheduler to
collect enough workgroups to launch the non-cooperative kernel, and the
non-cooperative kernel execution time. We make sure all kernels perform
their computations by comparing their result with expected results. We
had to ignore some configurations where the non-cooperative kernel asks
for all but one workgroups (namely bc, bfs and l-sssp on all chips, and
color on Iris), and the game application on HD520, since they lead to
system issues like GPU deadlock or blue screens---which we tend to think
are manifestations of driver bugs.

@@

\paragraph{Results}  \ADComment{Discuss the results.}

\myfig~\ref{fig:fine-grained-timing} reports the measure for three
configurations and represents the gather and execution time of the
non-cooperative kernel w.r.t. the quantity of workgroups allocated to
it. Times are in milliseconds, and a logarithmic scale is used since
gather times tend to be much smaller than execution times. The
horizontal grey lines indicates the desired rate for non-cooperative
kernels.

Responsiveness of the cooperative scheduling is measured by the gather
time. In all examples, we observe that the gather time grows as we ask
for more workgroups for the non-cooperative kernel: this is expected,
since the scheduler must wait for more workgroups to call $\offerkill$
or to reach a resizing barrier. Still, notice that the query barrier
leads to better and stable gather times for graph applications, since it
can collect all needed workgroups at the first time they pass through
the barrier. Gather time also application-dependent since it is related
to how often the cooperative kernel workgroups hit $\offerkill$ or
$\resizingglobalbarrier$.

\HEComment{missing results for all but one in bfs, discussed in expe
setup, do we need to say it here ?}

@@

\HEComment{Figure gather time: replace ``O'' by ``M'' , ie max nb of active workgroups}

\HEComment{Figure gather time: add ``ms'' to time coordinate}

\HEComment{Figure gather time: lower legend in hd5500 bfs figure}

\HEComment{Figure gather time: use consistent line styles for work stealing (ie, same style for ``heavy'' across all figures)}

>>>>

@@

\ADComment{This could be shorter.}
However, these are not hard-and-fast rules: a scheduler might employ
heuristics based on historical workloads to avoid thrashing.  For
example, if graphics calls are known to arrive at a certain rate then
the scheduler might decide that it is worth accepting an $\offerkill$
from the cooperative kernel on the assumption that a graphics task
will be requested momentarily.  Similarly, the scheduler might provide
some, but not all, compute units to the cooperative kernel in response
to an $\offerfork$ call if it has reason to believe that the remaining
compute units are likely to be useful for other high priority tasks
that are likely to arrive imminently.  With some additional language extensions, it would be possible for the cooperative kernel to hint to the scheduler if, for example, the kernel is at a point of computation where there is less work to do than usual, so that $\offerkill$ calls could be accepted in order to save energy.

@@

\ADComment{One cooperative, one non-cooperative.  Justification:
  sufficient to evaluate the key use case of a long-running irregular
  algorithm co-existing with a series of short-running tasks.
  Simplifieid task of implementing prototype, which we do not want to
  over-enginneer since it is only a mock-up.}

\ADComment{Restrictions on kernel structure: variables at top of
  kernel are implicitly $\transmit$-annotated, and $\offerkill$,
  $\offerfork$ and $\resizingglobalbarrier$ calls must appear only in
  kernel entry function.  Avoiding $\transmit$ allowed us to reuse the
  clang OpenCL front-end, and restricitng starting points for fork to
  be in main kernel allowed a ``goto''-based approach to redirecting
  control flow; threading this through procedure calls would be
  possible but laborious.  Restricting kill points enables kill to be
  implemented as ``return''; exiting from a procedure call stack would
  require stack unwinding support, akin to throwing an exception.
  Again make point that all these issues are only relevant due to our
  mock-up.}

@@

\ADComment{Explain kernel merge tool briefly.  Explain library
  briefly.}

\TSComment{we should mention that non-cooperative kernels must be
written in a persistent thread style}

@@

\paragraph{Safe implementation}

Concurrent calls to $\offerfork()$ must be protected with a mutex,
otherwise workgroups calling $\offerfork()$ will conflict over the
available workgroups. The same conflict over available workgroups can
occur at a new kernel launch between the scheduler and a workgroup
calling $\offerfork()$ occurs between the scheduler assigning tasks.
Thus we provide a mutex, called the \emph{assignment mutex}, in the
scheduler context which must be held anytime a group of available
resources is assigned to a kernel.

The provides an indirect mutex protecting concurrent calls to
$\offerfork()$ and $\offerkill()$. That is, $\offerkill()$ only
commands a workgroup to exit when the scheduler holds the
\emph{assignment lock}, as the scheduler is trying to obtain work
groups to send to a task, and $\offerfork()$ cannot spawn workgroups
without holding this lock.

The semantics of $\offerkill()$, which allow only the workgroup with
the highest id provide natural mutual exclusion for exiting workgroups
(as part of the exit, the kernel context is updated). That is, a
workgroup can only exit at $\offerkill()$ if it is the workgroup with
the highest id. An exiting thread releases this natural mutex by
decrementing the number of active workgroups in the kernel context.

@@

In section~\ref{??}, we show that a resizing barrier is semantically
equivalent to three normal barrier calls with calls to cfork and ckill
in between. It is possible, however to implement the resizing with
only one call to a global barrier with $\offerfork$ and $\offerkill$
baked into the implementation. \TSComment{note here that these
  implementations require no cfork or ckill outside of the barrier}

We consider barriers which have the master/slave model
(e.g. \cite{??}). That is, one workgroup (the master) collects signals
from the other workgroups (the slaves) indicating that they have
arrived at the barrier and are waiting for a reply indicating that
they may leave the barrier. Once the master has recieved a signal from
all the slaves, it replies with a signal saying that the slaves may
leave. \TSComment{Need to discuss new active workgroup number that
is stable across iterations}

A first-attempt at incorporating $\offerfork$ and $\offerkill$ into
such a barrier is straight forward. Upon entering the barrier, the
slaves first execute $\offerkill$, possibly exiting. The master then
waits for $M$ slaves (the number of active workgroups), which may
decrease, (based on the $\offerkill$ instructions executed by the
slaves), but will not increase. Once the master observes $M$ slaves
have arrived, it knows all other workgroups are in the kernel are
spinning, waiting to be released. The master executes $\offerfork$,
where there is a conditional for the potentially newly spawned threads
to join the other slaves waiting to be released. Finally, the master
releases all the slaves: the original slaves and the new slaves that
joined at $\offerfork$.

While this barrier implementation works, it has its flaws. Workgroups
only execute $\offerkill$ once per global barrier call. Depending on
the order in which workgroups arrive at the barrier, it is possible
that only one workgroup is killed per barrier call which could lead to
a high time for the scheduler to gather the sufficient amount of
workgroups to execute another kernel concurrently.

We observe if we provide an additional scheduler functionality for a
workgroup executing a cooperative kernel, we can greatly reduce the
gather time. We call the new function {\tt query()}. The semantics
simply return the number of workgroups that the scheduler requires to
kill from the cooperative kernel. A new barrier can now be implented
as follows: (1) Master waits for all slaves; (2) Master forks and sets
new workgroups as slaves. (3) Master calls {\tt query()} and gets
value $W$; (4) master releases slaves and broadcasts the value from
{\tt query} to slaves. (5) Workgroups with id greater than $M-W$ spin
on $\offerkill$, as they know from {\tt query()} that the scheduler
needs them to exit. We show in Section~\ref{} that the barrier using
query greatly reduces the gather time in practice.

@@

Our experiments span 13 pairs of cooperative kernels and data sets, 3
classes of non-cooperative kernel workloads and frequency, 4 quantities
of workgroups claimed for the non-cooperative kernel and 2 variations of
resizing barriers for graph algorithms, leading to 288
configurations. Each of this configuration is run 10 times on each
GPU. For each run, we record the execution time of the cooperative
kernel. For each scheduling of the non-cooperative kernel during the
run, we also record the \emph{gather time} needed by the scheduler to
collect enough workgroups to launch the non-cooperative kernel, and the
non-cooperative kernel execution time. We make sure all kernels perform
their computations by comparing their result with expected results. We
had to ignore some configurations where the non-cooperative kernel asks
for all but one workgroups (namely bc, bfs and l-sssp on all chips, and
color on Iris), and the game application on HD520, since they lead to
system issues like GPU deadlock or blue screens---which we think are
manifestations of driver bugs.

@@

\paragraph{Results}  \ADComment{Discuss the results.}
explain horizontal lines

missing results for all but one in bfs rmat22 since errors

also, high gather time corresponds to applications and data sets where
resize/kill is triggered less often.

Each configuration was run 10 times on each GPU.

@@
